Calendar-based graphics for visualizing people's daily schedules

### 2017-07-22 main.Rmd
changed to us english spelling

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organises and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly interval by sensors scattered around the city. Facetting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

keywords:
  # at least one keyword must be supplied
  formatted: [data visualisation, statistical graphics, time series, R package, grammar of graphics]
  plain:     [data visualisation, statistical graphics, time series, R package, grammar of graphics]
preamble: >
  \usepackage{amsmath}
  \usepackage{tikz}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area up until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of the foot traffic at 3 different sensors. Figure \ref{fig:facet-time} provides more details of the temporal patterns by faceting on day of the week. The sensor data, like many temporal datasets on human behaviours, lends itself to a number of exploratory data visualisation challenges:

@@

A collection of data plots, organised in a familiar format, offer a useful and intuitive way to understand complex data. 
@Wickham2012glyph embedded a sequence of daily temperature line charts into a glyph map according to the spatial locations; and @R-geofacet provided methods in the \pkg{geofacet} package to arrange data plots into a grid, while preserving the geographical layout. They both attempt to enable one to see the spatial context of each self-contained data plot.

@@

The remainder of the paper is organised as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in depth. Section \ref{sec:opt} lists and describes the options that come with the \code{frame_calendar} function. Section \ref{sec:examples} presents some variations of its usage. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

@@

The algorithm for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar, requires cells for days, embedded in blocks corresponding to months, organised into a grid layout for a year. Each month can be captured with 35 (5 $\times$ 7) cells, where the top left is Monday of week 1, and the bottom right is Sunday of week 5. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday-Sunday, which is determined by the year of the calendar. Months are of different length days, ranging from 28-31, and each month could extend over six weeks but the convention in these months is to wrap the last few days up to the top row of the block. The notation for creating these cells is as follows:

@@

The algorithm can be extended to display data from a single month up to a few years. The number of rows and columns in the layout can be specified with the arguments \code{nrow} and \code{ncol}. If the one would like to visualise data spanning over three years, for example, \code{nrow = 12} and \code{ncol = 3} would be an appropriate choice when assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid out horizontally. The vertical direction can be enabled by swapping $i$ and $j$ in equation \ref{eq:grid}, which occurs when the argument \code{dir} is set to \code{"v"}. This benefits users who are accustomed to calendars that are organised vertically, which is common in some countries. Next we shall describe some of the arguments that allow for alternate displays.

@@

```{r flinders-polar, fig.height = 6.5, fig.cap = "Star plots of hourly pedestrian counts at Flinders Street Station, which are line charts placed in polar coordinates. The periodic behaviours on workdays are clearly visible."}
```

@@

The \code{frame_calendar} function does not constrain itself to mapping only the temporal variable to the \code{x} argument. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organised in the daily calendar layout. It provides a visual tool for identifying repeating patterns. Figure \ref{fig:scatterplot} indicates two separate paths in the work-day glyphs, suggesting that an hour of a work day with many pedestrians is likely to be followed in the next hour with either greater or fewer numbers of pedestrians. However, this pattern is not present on non-work days. There is a bimodality on work days while there is unimodality on the rest of the days, which is also supported by Figure \ref{fig:flinders-2016}.

@@

The calendar-based visualisation provides data plots in the familiar (at least for the Western world) format of an everyday tool. Special events for the region, like Anzac Day in Australia, or Thanksgiving Day in the USA, more easily pop out to the viewer as public holidays, rather than a typical work day.

This sort of layout may be useful for studying consumer trends, or human behaviour, like the pedestrian patterns. It may not work so well for physical patterns like temperature, which are not typically affected by human activity.

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualizing people's daily schedules"
  plain: "Calendar-based graphics for visualizing people's daily schedules"
abstract: >
  This paper describes a \code{frame\_calendar} function that organizes and displays temporal data, collected on sub-daily resolution, into a calendar layout. Calendars are broadly used in society to display temporal information, and events. The \code{frame\_calendar} utilizes linear algebra on the date  variable to create the layout. It utilizes the grammar of graphics to create the plots inside each cell, and thus synchronises neatly with \pkg{ggplot2} graphics. The motivating application is studying pedestrian behavior, based on counts which are captured at hourly interval by sensors scattered around the city. Facetting by the usual features such as day and month, was insufficient to examine the behavior. Making displays on a monthly calendar format helps to understand pedestrian patterns relative to  events such as work days, weekends, holidays, and special events. The layout algorithm has several format options and variations. It is implemented in the R package \pkg{sugrrants}.

keywords:
  # at least one keyword must be supplied
  formatted: [data visualization, statistical graphics, time series, R package, grammar of graphics]
  plain:     [data visualization, statistical graphics, time series, R package, grammar of graphics]
preamble: >
  \usepackage{amsmath}
  \usepackage{tikz}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 sensors installed that count pedestrians every hour across the downtown area up until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to catch a glimpse of such data. A small multiples, shown in Figure \ref{fig:time-series-plot}, gives an overall picture of the foot traffic at 3 different sensors. Figure \ref{fig:facet-time} provides more details of the temporal patterns by faceting on day of the week. The sensor data, like many temporal datasets on human behaviors, lends itself to a number of exploratory data visualization challenges:

@@

A collection of data plots, organized in a familiar format, offer a useful and intuitive way to understand complex data. 
@Wickham2012glyph embedded a sequence of daily temperature line charts into a glyph map according to the spatial locations; and @R-geofacet provided methods in the \pkg{geofacet} package to arrange data plots into a grid, while preserving the geographical layout. They both attempt to enable one to see the spatial context of each self-contained data plot.

@@

The remainder of the paper is organized as follows. Section \ref{sec:algorithm} demonstrates the construction of the calendar layout in depth. Section \ref{sec:opt} lists and describes the options that come with the \code{frame_calendar} function. Section \ref{sec:examples} presents some variations of its usage. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

@@

The algorithm for constructing a calendar plot uses linear algebra, similar to that used in the glyph map displays for spatio-temporal data [@Wickham2012glyph]. To make a year long calendar, requires cells for days, embedded in blocks corresponding to months, organized into a grid layout for a year. Each month can be captured with 35 (5 $\times$ 7) cells, where the top left is Monday of week 1, and the bottom right is Sunday of week 5. These cells provide a micro canvas on which to plot the data. The first day of the month could be any of Monday-Sunday, which is determined by the year of the calendar. Months are of different length days, ranging from 28-31, and each month could extend over six weeks but the convention in these months is to wrap the last few days up to the top row of the block. The notation for creating these cells is as follows:

@@

The algorithm can be extended to display data from a single month up to a few years. The number of rows and columns in the layout can be specified with the arguments \code{nrow} and \code{ncol}. If the one would like to visualize data spanning over three years, for example, \code{nrow = 12} and \code{ncol = 3} would be an appropriate choice when assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid out horizontally. The vertical direction can be enabled by swapping $i$ and $j$ in equation \ref{eq:grid}, which occurs when the argument \code{dir} is set to \code{"v"}. This benefits users who are accustomed to calendars that are organized vertically, which is common in some countries. Next we shall describe some of the arguments that allow for alternate displays.

@@

```{r flinders-polar, fig.height = 6.5, fig.cap = "Star plots of hourly pedestrian counts at Flinders Street Station, which are line charts placed in polar coordinates. The periodic behaviors on workdays are clearly visible."}
```

@@

The \code{frame_calendar} function does not constrain itself to mapping only the temporal variable to the \code{x} argument. Figure \ref{fig:scatterplot} shows a lag scatterplot at Flinders Street Station, where the current hourly count is assigned to the \code{x} argument and the lagged hourly count to the \code{y} argument. This figure is organized in the daily calendar layout. It provides a visual tool for identifying repeating patterns. Figure \ref{fig:scatterplot} indicates two separate paths in the work-day glyphs, suggesting that an hour of a work day with many pedestrians is likely to be followed in the next hour with either greater or fewer numbers of pedestrians. However, this pattern is not present on non-work days. There is a bimodality on work days while there is unimodality on the rest of the days, which is also supported by Figure \ref{fig:flinders-2016}.

@@

The calendar-based visualization provides data plots in the familiar (at least for the Western world) format of an everyday tool. Special events for the region, like Anzac Day in Australia, or Thanksgiving Day in the USA, more easily pop out to the viewer as public holidays, rather than a typical work day.

This sort of layout may be useful for studying consumer trends, or human behavior, like the pedestrian patterns. It may not work so well for physical patterns like temperature, which are not typically affected by human activity.

### 2017-07-22 main.tex
changed to us english spelling

\author{
Earo Wang\\Monash University \And Dianne Cook\\Monash University \And Rob J Hyndman\\Monash University
}
\title{Calendar-based graphics for visualising people's daily schedules}

\Plainauthor{Earo Wang, Dianne Cook, Rob J Hyndman}
\Plaintitle{Calendar-based graphics for visualising people's daily schedules}

\Abstract{
This paper describes a \code{frame\_calendar} function that organises
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly interval by sensors scattered around the city.
Facetting by the usual features such as day and month, was insufficient
to examine the behavior. Making displays on a monthly calendar format
helps to understand pedestrian patterns relative to events such as work
days, weekends, holidays, and special events. The layout algorithm has
several format options and variations. It is implemented in the R
package \pkg{sugrrants}.
}

\Keywords{data visualisation, statistical graphics, time series, R package, grammar of graphics}
\Plainkeywords{data visualisation, statistical graphics, time series, R package, grammar of graphics}

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 sensors installed that
count pedestrians every hour across the downtown area up until the end
of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. A small multiples, shown in Figure
\ref{fig:time-series-plot}, gives an overall picture of the foot traffic
at 3 different sensors. Figure \ref{fig:facet-time} provides more
details of the temporal patterns by faceting on day of the week. The
sensor data, like many temporal datasets on human behaviours, lends
itself to a number of exploratory data visualisation challenges:

@@

A collection of data plots, organised in a familiar format, offer a
useful and intuitive way to understand complex data.
\citet{Wickham2012glyph} embedded a sequence of daily temperature line
charts into a glyph map according to the spatial locations; and
\citet{R-geofacet} provided methods in the \pkg{geofacet} package to
arrange data plots into a grid, while preserving the geographical
layout. They both attempt to enable one to see the spatial context of
each self-contained data plot.

@@

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
\citep{Wickham2012glyph}. To make a year long calendar, requires cells
for days, embedded in blocks corresponding to months, organised into a
grid layout for a year. Each month can be captured with 35 (5 \(\times\)
7) cells, where the top left is Monday of week 1, and the bottom right
is Sunday of week 5. These cells provide a micro canvas on which to plot
the data. The first day of the month could be any of Monday-Sunday,
which is determined by the year of the calendar. Months are of different
length days, ranging from 28-31, and each month could extend over six
weeks but the convention in these months is to wrap the last few days up
to the top row of the block. The notation for creating these cells is as
follows:

@@

The algorithm can be extended to display data from a single month up to
a few years. The number of rows and columns in the layout can be
specified with the arguments \code{nrow} and \code{ncol}. If the one
would like to visualise data spanning over three years, for example,
\code{nrow = 12} and \code{ncol = 3} would be an appropriate choice when
assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid
out horizontally. The vertical direction can be enabled by swapping
\(i\) and \(j\) in equation \ref{eq:grid}, which occurs when the
argument \code{dir} is set to \code{"v"}. This benefits users who are
accustomed to calendars that are organised vertically, which is common
in some countries. Next we shall describe some of the arguments that
allow for alternate displays.

@@

\caption[Star plots of hourly pedestrian counts at Flinders Street Station, which are line charts placed in polar coordinates]{Star plots of hourly pedestrian counts at Flinders Street Station, which are line charts placed in polar coordinates. The periodic behaviours on workdays are clearly visible.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}

>>>>

\author{
Earo Wang\\Monash University \And Dianne Cook\\Monash University \And Rob J Hyndman\\Monash University
}
\title{Calendar-based graphics for visualizing people's daily schedules}

\Plainauthor{Earo Wang, Dianne Cook, Rob J Hyndman}
\Plaintitle{Calendar-based graphics for visualizing people's daily schedules}

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} utilizes
linear algebra on the date variable to create the layout. It utilizes
the grammar of graphics to create the plots inside each cell, and thus
synchronises neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior, based on counts which are
captured at hourly interval by sensors scattered around the city.
Facetting by the usual features such as day and month, was insufficient
to examine the behavior. Making displays on a monthly calendar format
helps to understand pedestrian patterns relative to events such as work
days, weekends, holidays, and special events. The layout algorithm has
several format options and variations. It is implemented in the R
package \pkg{sugrrants}.
}

\Keywords{data visualization, statistical graphics, time series, R package, grammar of graphics}
\Plainkeywords{data visualization, statistical graphics, time series, R package, grammar of graphics}

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 sensors installed that
count pedestrians every hour across the downtown area up until the end
of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed light into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to catch a glimpse of such data. A small multiples, shown in Figure
\ref{fig:time-series-plot}, gives an overall picture of the foot traffic
at 3 different sensors. Figure \ref{fig:facet-time} provides more
details of the temporal patterns by faceting on day of the week. The
sensor data, like many temporal datasets on human behaviors, lends
itself to a number of exploratory data visualization challenges:

@@

A collection of data plots, organized in a familiar format, offer a
useful and intuitive way to understand complex data.
\citet{Wickham2012glyph} embedded a sequence of daily temperature line
charts into a glyph map according to the spatial locations; and
\citet{R-geofacet} provided methods in the \pkg{geofacet} package to
arrange data plots into a grid, while preserving the geographical
layout. They both attempt to enable one to see the spatial context of
each self-contained data plot.

@@

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
\citep{Wickham2012glyph}. To make a year long calendar, requires cells
for days, embedded in blocks corresponding to months, organized into a
grid layout for a year. Each month can be captured with 35 (5 \(\times\)
7) cells, where the top left is Monday of week 1, and the bottom right
is Sunday of week 5. These cells provide a micro canvas on which to plot
the data. The first day of the month could be any of Monday-Sunday,
which is determined by the year of the calendar. Months are of different
length days, ranging from 28-31, and each month could extend over six
weeks but the convention in these months is to wrap the last few days up
to the top row of the block. The notation for creating these cells is as
follows:

@@

The algorithm can be extended to display data from a single month up to
a few years. The number of rows and columns in the layout can be
specified with the arguments \code{nrow} and \code{ncol}. If the one
would like to visualize data spanning over three years, for example,
\code{nrow = 12} and \code{ncol = 3} would be an appropriate choice when
assessing the differences of a given month across the years.

In Section \ref{sec:algorithm} we only illustrated that grids are laid
out horizontally. The vertical direction can be enabled by swapping
\(i\) and \(j\) in equation \ref{eq:grid}, which occurs when the
argument \code{dir} is set to \code{"v"}. This benefits users who are
accustomed to calendars that are organized vertically, which is common
in some countries. Next we shall describe some of the arguments that
allow for alternate displays.

@@

\caption[Star plots of hourly pedestrian counts at Flinders Street Station, which are line charts placed in polar coordinates]{Star plots of hourly pedestrian counts at Flinders Street Station, which are line charts placed in polar coordinates. The periodic behaviors on workdays are clearly visible.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}
